Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-04-01 paper/paper1/ms.tex
Writing prose in the discussion sections

Tens of thousands or more individual spectral lines give rise to a sea of undulations that imbue each stellar spectrum with its characteristic appearance.  The identification and understanding of these lines has defined a large category of astrophysics over the last century.  The field grew from by-eye catalogs of stellar templates \citep{1901AnHar..28..129C} to quantifying the role of atomic ionization balance \citep{1925PhDT.........1P}, to modern synthetic forward models including millions or billions of lines \citep{husser13, 2021ApJ...920...85M}.  As technology has improved, our data and models have become more voluminous, more precise, and more complicated.  The mere act of comparing models to observed spectra can resemble a computational statistics challenge as much as a scientific one.  Here we introduce a machine-learning-based framework \texttt{blas\'e} aimed at solving computational, statistical, and scientific challenges associated with data-model comparisons for modern astronomical spectroscopy.

@@

In this paper we show that autodiff-aware semi-empirical models offer an appealing middle ground: informed from self-consistent models, but refined with data.  In Section \ref{methodology} we introduce the \emph{blas\'e} forward-model design and its \texttt{PyTorch}-based \cite{2019arXiv191201703P} implementation.  We show how to supplant the computationally expensive radiative transfer by cloning pre-computed synthetic spectral models.  In Section \ref{transferLearn} we describe how to fine-adjust the cloned models to real-world data from high-grasp near-IR spectrographs, using a transfer-learning step.  Section \ref{secInjRec} exhibits injection-recovery tests on simulated data of known properties, revealing the scalability and fundamental limits of the technique.  Finally we discuss many conceivable extensions to the framework, and directions for future work in Section \ref{secDiscuss}.

@@

\begin{eqnarray}
    \mathcal{L} = \sum_i^{N_S} (S_i - S_{\mathrm{clone},i})^2 = \mathsf{R^\intercal}\cdot \mathsf{R}
\end{eqnarray}

@@

\subsection{Performance of transfer learning}

\begin{outline}
    \1 What is the residual level with bare PHOENIX (i.e. tuning continuum only, not lines)?
    \1 What is the residual level with no regularization (i.e. tune all lines and continuum)?
    \2 Should be near-zero except for missing lines
    \2 What should we do about missing lines?
    \1 What is the residual level with modest regularization? What is the typical change to the line?
    \2 Plot of cloned FWHM versus FWHM transferred
    \2 The pseudo-Voigt line properties can be analytically integrated to give an equivalent width
    \2 Plot of cloned EW (before) versus transfer EW (after)
\end{outline}

\section{Injection/recovery tests}\label{secInjRec}
Here we conduct injection/recovery tests.

@@

@@

The empirical version, \texttt{specmatch-emp} \citep{2017ApJ...836...77Y} matched spectra better than the synthetic templates, but is still too rigid for some applications and requires the assembly of hundreds of standardized high signal-to-noise-ratio templates, ideally with low intrinsic rotational broadening.  Such a large number of high-quality templates has not yet been established in the near infrared, and may not be feasible for brown dwarfs.

@@

\begin{outline}
    \1 We see some lines that devolve into missing continuum opacity
    \1 How do the lineshapes cluster in $\sigma_G$ vs $FWHM_L$?
    \1 How do the lineshapes cluster in Amplitude vs FWHM? Can some lines be rejected?
\end{outline}

\subsection{Alternatives to the PseudoVoigt Profile}
Two sampled Voigt approaches could be feasible. Direct convolution of a sampled Gaussian and Lorentzian profile may be too costly and inexact if pixels are coarsely sampled. Analogously, the convolution can be achieved in the Fourier domain through the multiplication of the Fast Fourier Transforms of the Gaussian and Lorentzian profiles, followed by the subsequent Inverse Fourier Transforms. The main rationale for either of these two choices would be to anticipate and align with planned convolutions such as rotational broadening and instrumental kernels. We do not currently implement these sampled convolutions in \emph{blas\'e} due to their prohibitive computational cost.

\subsection{Prospects for closed-form computation of Jacobians}
In principle, we could compute the Jacobians analytically---since the partial derivatives of the Voigt function are known in closed-form---circumventing the vendor lock-in and memory overhead of an autodiff implementation with PyTorch. Further, an autodiff-aware implementation of the \emph{celerit\'e} algorithm \citep{2017AJ....154..220F} does not yet exist in PyTorch, constricting our choices for simultaneous continuum fitting. In practice, PyTorch also offers hardware acceleration on GPU or TPU \citep{10.1145/3079856.3080246}, mature support for a range of optimizers, and other perks that motivated its choice over analytic implementations or other machine learning frameworks. The GPyTorch framework \citep{2018arXiv180911165G} may offer a workaround to the absence of a \emph{celerit\'e} implementation. Still, the performance of an analytic Jacobian implementation over the autodiff route presented here offers an interesting avenue of future research.

\subsection{Tuning the continuum fit}
We could also tune or freeze an experimental wavelength-dependent pre-factor to Equation \ref{equation1} to correct for lingering imperfections in the otherwise-fixed continuum flattening procedure.

>>>>

Tens of thousands or more individual spectral lines give rise to a sea of undulations that imbue each stellar spectrum with its characteristic appearance.  The identification and understanding of these lines has defined a large category of astrophysics over the last century.  The field grew from by-eye catalogs of stellar templates \citep{1901AnHar..28..129C} to quantifying the role of atomic ionization balance \citep{1925PhDT.........1P}, to modern synthetic forward models including millions or billions of lines \citep{husser13, 2021ApJ...920...85M}.  As technology has improved, our data and models have become more voluminous, more precise, and more complicated.  The mere act of comparing models to observed spectra can now resemble a computational statistics challenge as much as a scientific one.  Here we introduce a new machine-learning-based framework \texttt{blas\'e} aimed at solving computational, statistical, and scientific challenges associated with data-model comparisons for modern astronomical spectroscopy.

@@

In this paper we show that autodiff-aware semi-empirical models offer an appealing middle ground: informed from self-consistent models, but refined with data.  In Section \ref{methodology} we introduce the \emph{blas\'e} forward-model design and its \texttt{PyTorch}-based \cite{2019arXiv191201703P} implementation.  We show how to supplant the computationally expensive radiative transfer by cloning pre-computed synthetic spectral models.  We show how to combine both stellar and telluric models in a joint modelling approach.  In Section \ref{transferLearn} we describe how to fine-adjust the cloned models to real-world data from high-grasp near-IR spectrographs, using a transfer-learning step.
%Section \ref{secInjRec} exhibits injection-recovery tests on simulated data of known properties, revealing the scalability and fundamental limits of the technique.  
Finally we discuss many conceivable extensions to the framework, and directions for future work in Section \ref{secDiscuss}.

@@

\begin{eqnarray}
    \mathcal{L} = \sum_i^{N_S} (S_i - S_{\mathrm{clone},i})^2 = \mathsf{R^\intercal}\cdot \mathsf{R} \label{simpleLikelihood}
\end{eqnarray}

@@

% \begin{outline}
%     \1 What is the residual level with bare PHOENIX (i.e. tuning continuum only, not lines)?
%     \1 What is the residual level with no regularization (i.e. tune all lines and continuum)?
%     \2 Should be near-zero except for missing lines
%     \2 What should we do about missing lines?
%     \1 What is the residual level with modest regularization? What is the typical change to the line?
%     \2 Plot of cloned FWHM versus FWHM transferred
%     \2 The pseudo-Voigt line properties can be analytically integrated to give an equivalent width
%     \2 Plot of cloned EW (before) versus transfer EW (after)
% \end{outline}

%\section{Injection/recovery tests}\label{secInjRec}
%Here we conduct injection/recovery tests.

@@

Several astronomical spectral frameworks share similar aims as \texttt{blas\'e}.  These existing frameworks will have enduring value for the wide range of problems in the field of stellar spectroscopy.  Here we scrutinize the differences among some of these approaches to clarify how This Work fits in.

@@

The empirical version, \texttt{specmatch-emp} \citep{2017ApJ...836...77Y} matched spectra better than the synthetic templates, but is still too rigid for some applications and requires the assembly of hundreds of standardized high signal-to-noise-ratio templates, ideally with low intrinsic rotational broadening.  Such a large number of high-quality templates with high resolving power and low $v\sin{i}$ has not yet been established in the near infrared.

@@

\section{Conceivable Improvements, Extensions, and Limitations}

As currently implemented, \texttt{blas\'e} represents a minimum viable product: it out-performs some existing approaches for some science cases.  The scaffolding of the framework is designed in a way to promote extensibility, so the future of the framework is bright.  Here we enumerate some planned or conceivable extensions to the framework that would unlock new science arenas.  We order the presentation of these extensions from our perceived ease-of-implementation, from easiest to hardest.

\subsection{Exact instead of pseudo Voigt Profile}
We currently employ the pseudo-Voigt profile for its low computational cost.  We have a prototype exact-Voigt-Hjertingimplementation following \citet{2022ApJS..258...31K}.  We coarsely estimate that moving to this exact-Voigt implementation wculd decrease some residual regions by $\sim 30\%$, while increasing the computational cost by more than $10\times$ over the existing pseudo-Voigt approximation.  The exact-Voigt-Hjerting implementation still outbids the higher cost of a direct numerical convolution of a Gaussian and Lorentzian profile.

\subsection{Prospects for closed-form computation of Jacobians}
In principle, one could compute the Jacobians analytically---since the partial derivatives of the Voigt function are known in closed-form.  This analytical approach would cut-out the autodiff approach, using the exact Jacobians to compute the Gradient Descent step.  The analytical approach may be faster and/or possess lower memory overhead than the autodiff one.  However, working out the chain-rule for the closed-form analytical Jacobians may become difficult or impossible once combined with the many augmentations get applied at the joint telluric-and-stellar modeling step.  So analytic Jacobians could principally accelerate the cloning step, but would add significant implementation complexity at the important data-model comparison step.  We therefore deem this extension somewhat diminishing returns on scarce developer time.

\subsection{Addressing the pseudo-continuum with Gaussian Process regression}

We currently assume the input spectra are perfectly normalized to the continuum.  This approximation impacts all training stages, including the stellar cloning, telluric cloning, and data-model comparison.  Gaussian Processes (GPs) could and should be brought to bear at each of these optimization stages, as described in the ``global kernels'' in \texttt{Starfish} \citet{czekala15}.  In short, a GP-likelihood relaxes the assumption that the continuum has been perfectly normalized, in favor of the more realistic statement ``the continuum has been coarsely normalized, with some characteristic-but-as-yet-unknown correlation and scale length and amplitude of the imperfections''.  That statement translated to the following modification to Equation \ref{simpleLikelihood}:

\begin{eqnarray}
    \mathcal{L} =  \frac{1}{2}\mathsf{R^\intercal} \mathsf{C}^{-1} \mathsf{R} +\frac{1}{2}\ln{\det{\mathsf{C}}} \label{GPLikelihood}
\end{eqnarray}

where we introduce the covariance matrix $\mathsf{C}$, with its concommittant kernel and collection of typically 2-3 parameters.  We anticipate that this GP likelihood would have the greatest impact on stellar stars with significant band-heads and line-blanketing: spectra with a so-called ``pseudo-continuum''.  M-dwarfs and brown dwarfs fall into this fear-inducing category.

The main demerit of moving to a GP-likelihood is computational cost.  Fortunately a few efficient autodiff-aware implementations of GPs exist. The \texttt{celerite} algorithm \emph{celerit\'e} \citep{2017AJ....154..220F} has an exact backpropation implementation \citep{2018RNAAS...2...31F} that scales linearly with the number of data points.  The celerite algorithm does not currently have a PyTorch implementation.  The \texttt{GPyTorch} framework \citep{2018arXiv180911165G} has a large category of approximate and exact GPs that could be straight-forwardly dropped into \texttt{blas\'e}.  Even still these GPs could increase the computation cost by of order $10\times$.

An alternative to GPs could be to simply tune the $\mathsf{P}$ term that represents the wavelength-dependent pre-factor to Equation \ref{equation1}.  Tuning $\mathsf{P}$ would correct for only large-scale imperfections in the otherwise-fixed continuum flattening procedure.
